---
title: "IMOS_SRS_SST"
author: "Denisse Fierro Arcos"
date: "2025-07-08"
output: 
  github_document:
    toc: true
    html_preview: true
always_allow_html: true
---

```{r, echo = F}
knitr::opts_chunk$set(fig.path = "figures/") 
```


# Introduction

In this notebook, we will explore how to access and visualize the IMOS SRS Sea Surface Temperature (SST) data using R. We will also show how to extract SST data using a point of interest. 

This SST data product is available through the IMOS THREDDS Data Server and it provides the skin sea surface temperature which was measured using infra-red radiometers on various satellites. The data is composited over multiple swaths/scenes and gridded over a 0.02 degree rectangular grid over Australia. For more information, see http://www.ghrsst.org

# Loading required libraries
  
```{r libraries, results = "hide", message = F, warning = F}
#Connection to THREDDS
library(thredds)
#Downloading data
library(curl)
#Data manipulation
library(dplyr)
library(stringr)
library(purrr)
#Spatial data
library(sf)
#Gridded data
library(terra)
#Plotting
library(ggplot2)
library(tidyterra)
#Base map
library(rnaturalearth)
#Extracting data from IMOS
library(remora)
```

# Connecting to IMOS THREDDS server
The IMOS THREDDS server can be accessed using the `thredds` package in R. It is also possible to explore the catalog to find the dataset we are interested in, which we show below, but we recommend using the web browser for easier navigation. 
  
```{r connect_thredds}
# We define the URL of the IMOS THREDDS catalog
imos_cat_url <- "https://thredds.aodn.org.au/thredds/catalog/IMOS/catalog.xml"
# We create a CatalogNode object to connect to the IMOS THREDDS catalog
imos_cat <- CatalogNode$new(imos_cat_url, prefix = "thredds")
```

## Exploring the IMOS THREDDS catalog
We can open the catalog in the web browser to explore the datasets available as follows:
  
```{r}
imos_cat$browse()
```
  
Once in the browser, you can change the extension of the catalog URL from `catalog.xml` to `catalog.html`. This way you can explore the catalog in a more user-friendly format.  
  
As mentioned previously, we can also explore the catalog programmatically, but this is more time-consuming. We will check the SST datasets inside the SRS catalog:
  
```{r}
# Listing datasets in the catalog
imos_cat$get_catalogs()
```
  
We can see there are a number of datasets available in the IMOS catalog. We are interested in the `SRS` products, so we can check the datasets available in the `SRS` catalog:
  
```{r}
imos_cat$get_catalogs()[["SRS"]]
```
  
We can see that the `SRS` catalog contains the `SST` dataset we are interested in. There are multiple datasets available inside `SST`, but below we show you the path to the `ghrsst` dataset, which contains the Sea Surface Temperature data we want to access:

```{r}
imos_cat$get_catalogs()[["SRS"]]$get_catalogs()[["SST"]]$
  get_catalogs()[["ghrsst"]]$get_catalogs()[["L3SM-1d"]]$
  get_catalogs()[["day"]]$get_catalogs()[["2025"]]
```
  
We can see that there are 163 files available in the `2025` folder, which contains the SST data for that year. We also get a URL that will allow us to explore the catalog further. We will use this URL to download the SST data for the first week of 2025.

```{r}
# First we connect to the catalog
srs_cat <- CatalogNode$new(paste0("https://thredds.aodn.org.au/thredds/catalog",
                                  "/IMOS/SRS/SST/ghrsst/L3SM-1d/day/2025/", 
                                  "catalog.xml"),
                           prefix = "thredds")

# Now we can get the dataset names available in the catalog
sst_jan_names <- srs_cat$get_dataset_names() |>
  str_subset("2025010[1-7]")

# Checking we have selected the correct datasets
sst_jan_names
```
  
We have seven datasets available for the first week of January 2025, which is exactly what we wanted. Once again, this method to explore the catalog is more time-consuming, so unless you're familiar with the structure of the catalog, we highlight recommend you use the web browser instead.
  
We now need a complete URL to download the SST data as shown below.

```{r}
# Get base url for HTTPserver access
http_base <- srs_cat$list_services()$http[["base"]]

# Put together full url to download selected files
sst_jan_urls <- sst_jan_names |> 
  map_chr(\(x) paste0("https://thredds.aodn.org.au", http_base,
                      srs_cat$get_datasets()[[x]]$get_url()))

# Checking results
sst_jan_urls
```
  
We are now ready to download the data.  

## Download SST data from IMOS THREDDS server
  
```{r, eval = FALSE}
# We will ensure we have a directory to store the data
if(!dir.exists("../data")){
  dir.create("../data", recursive = T)
}

# We will download the SST data for the first week of January 2025
# The file names will be named the same as in the catalog
map2(sst_jan_urls, sst_jan_names, 
     \(x, y) curl_download(x, file.path("../data/", y)))
```
  
Data download may take a few minutes depending on your internet connection. Once the data is downloaded, we can load it into R for further analysis.
  
# Load SST data
  
```{r}
# We can get a list of the downloaded files
sst_jan_files <- list.files("../data", pattern = "GHRSST", full.names = T)

# We can load the SST data using the terra package
sst_jan_w1 <- rast(sst_jan_files)

# Exploring SST data
sst_jan_w1
```
  
The dataset contains multiple variables, but we are interested in the `sea_surface_temperature` variable. We can subset the data to keep SST data only and then we will convert it from Kelvin to Celsius.

You may also have noticed that there is no information about dates in the raster object. We can extract this information from the `source` attribute of the raster object.

```{r}
# Subsetting SST data
sst_jan_w1 <- sst_jan_w1[[names(sst_jan_w1) == "sea_surface_temperature"]]

# Getting dates from the source attribute
date_sst_jan <- sources(sst_jan_w1) |> 
  str_extract(pattern = "2025010[1-7]") |> 
  as.Date(format = "%Y%m%d")

# Adding dates as time and names to the raster object
time(sst_jan_w1) <- date_sst_jan
names(sst_jan_w1) <- date_sst_jan

# Transforming from Kelvin to Celsius
sst_jan_w1_celsius <- sst_jan_w1 - 273.15
sst_jan_w1_celsius
```
  
# Extracting SST data for points of interest
We can extract SST data for specific points of interest using the `extract` function from the `terra` package. For example, we can extract SST data for a points in the east and west coast of Australia.

```{r}
# Defining points of interest
points_of_interest <- data.frame(lon = c(153.4, 115.7),
                                 lat = c(-19.5, -32)) |> 
  # Converting points to a spatial object
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# Extracting SST data for points of interest
sst_points <- extract(sst_jan_w1_celsius, points_of_interest, xy = T)

# Checking extracted SST data
sst_points
```
  
We can reformat the extracted data to make it easier to read and visualize.
  
```{r}
sst_points <- sst_points |> 
  pivot_longer(cols = !c(ID, x, y), names_to = "date",
               values_to = "sst_celsius")

sst_points
```
  
# Extracting SST data with `remora`
We can also use the `remora` package to extract SST data for our points of interest. This package provides a convenient way to access IMOS data for specific locations. More information about `remora` can be found in their website: https://imos-animaltracking.github.io/remora/.    
  
For this example, we will use the same points of interest defined above.  
  
```{r}
# We can explore the environment variables available in the remora package
imos_variables()
```
  
We will use `rs_sst` which is the same product we accessed via the THREDDS server. Note that this step may take a few minutes to complete.
  
```{r, eval = F}
remora_sst <- extractEnv(sst_points, X = "x", Y = "y", datetime = "date",
                         env_var = "rs_sst")
```
  
# Calculating mean SST for the first week of January 2025

```{r}
mean_sst_jan_w1 <- mean(sst_jan_w1_celsius, na.rm = TRUE)
mean_sst_jan_w1
```
  
# Plotting SST data
We can create a simple plot of the mean SST using base R.

```{r}
plot(mean_sst_jan_w1)
```
  
We can also create a more informative plot using `ggplot2` and `tidyterra` packages. We will use the `geom_spatraster` function to plot the raster data and add a base map of Australia using the `rnaturalearth` package.

```{r}
# Getting a base map of Australia
aus <- ne_countries(country = "Australia", scale = "medium",
                    returnclass = "sf")
```

Remember that when using `ggplot2`, we create a plot by stacking layers. This means we will need to add the raster data as the first layer and then add the base map as a second layer. We can also add points for the locations we extracted SST data from.  
  
```{r}
ggplot()+
  # Using tidyterra to plot the raster data
  geom_spatraster(data = mean_sst_jan_w1) +
  # Setting the color scale for the SST data
  scale_fill_viridis_c(option = "C", name = "SST (°C)") +
  # Adding base map of Australia
  geom_sf(data = aus, fill = NA, color = "grey") +
  # Adding points for the locations we extracted SST data from
  geom_sf(data = points_of_interest, color = "darkgreen", size = 3) +
  # Adding labels and theme
  labs(title = "Mean Sea Surface Temperature (SST) for January 2025",
       x = "Longitude", y = "Latitude") +
  theme_bw()
```
  
We can also make a plot comparing the mean SST for the first week of January 2025 at our two points of interest.

```{r}
sst_points |> 
  ggplot(aes(x = date, y = sst_celsius, color = as.factor(ID))) +
  geom_line(aes(group = ID)) +
  geom_point() +
  labs(title = "SST at Points of Interest for January 2025",
       x = "Date", y = "SST (°C)", color = "Point ID") +
  theme_bw()
```
  
Finally, we can also create a multipanel plot to visualize the SST data for each day of the first week of January 2025 at our points of interest.
  
```{r}
ggplot()+
  geom_spatraster(data = sst_jan_w1_celsius)+
  scale_fill_viridis_c(option = "C", name = "SST (°C)") +
  geom_sf(data = aus, fill = "#fff2cc", color = "#fff2cc") +
  facet_wrap(~lyr) +
  theme_bw() +
  labs(title = "Sea Surface Temperature (SST) for January 2025") +
  theme(legend.position = "bottom", legend.title.position = "top", 
        axis.text.x = element_text(angle = 45, hjust = 1))
```
  
Remember that you can save any plots as follows:
  
```{r, eval = FALSE}
ggsave("../outputs/sst_australia_jan2025.png", dpi = 300, 
       bg = "white")
```

# Conclusion
In this notebook, we have gone through the basics of how to access IMOS data available via the THREDDS server, specifically the Sea Surface Temperature (SST) data. We have shown how to download the data, load it into R, extract SST data for specific points of interest, and visualize the data using `ggplot2` and `tidyterra`. We have also shown how to calculate the mean SST for the first week of January 2025 and created multifaceted plots to visualise the data.

  