---
title: "Dugong aerial survey
author: "Denisse Fierro Arcos"
date: "2026-08-28"
output: 
  github_document:
    toc: true
    html_preview: true
always_allow_html: true
---

```{r, echo = F}
knitr::opts_chunk$set(fig.path = "figures/") 
```


# Introduction

This notebook shows how to access the `Dugong aerial survey observations in Northern Australia`. This dataset contains data obtained from aerial surveys that have been carried out by James Cook University from 1984 in nine regions across Northern Australia. 

This dataset is available as a parquet file, which is a file format designed to be efficient in storing and retrieving tabular data. Parquet files can be loaded as data frames as you would with other tabular data format like csv. The parquet files for the Dugong datasets can be accessed from the [AODN's Amazon server](https://registry.opendata.aws/). There are several other ways to access this dataset, which are described in the [IMOS Metadata Catalogue](https://catalogue-imos.aodn.org.au/geonetwork/srv/eng/catalog.search#/metadata/b85f6b58-636a-4a6e-aca4-c90999d7f424).  

# Loading required libraries
  
```{r libraries, results = "hide", message = F, warning = F}
library(arrow)
#Data manipulation
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(stringr)
#Plotting
library(ggplot2)
#Base maps
library(rnaturalearth)
#Spatial data
library(sf)
#Gridded data
library(terra)
#Connection to THREDDS
library(thredds)
#Downloading data
library(curl)
```

## Accessing the Dugong Aerial Survey dataset with `arrow`
From the metadata catalogue we obtain the URI and create a connection to the dataset using the `arrow` package. By using the `open_dataset` function, we are able to access the data stored in the server, but we don't load the dataset just yet. Instead we will inspect the contents of this dataset to check if our computer can handle the entire dataset being loaded.

```{r}
ds <- open_dataset(
  "s3://data-uplift-public/stored/datauplift/dugong/dugong.parquet")
```

We can check the file structure (i.e., schema) of the seabird dataset. This will give us the name of the columns contained in the dataset, which can help us filter the data before accessing.  

```{r}
ds$schema
```

From here, we can see there are `r length(ds$schema)` columns, but we can check how many records in total are included in the dataset as follows:  

```{r}
dim(ds)
```

This dataset has `r nrow(ds)` rows, which most computers are able to handle with ease, so we can load the entire dataset and explore its structure. However, `arrow` allows us to lazily load datasets, so if the dataset you are interested in has millions of rows, you can subset it or summarise the data and load into memory only what you need, instead of loading all records at once. This could be especially useful if you have limited computing resources, this way you can avoid overloading the computer's memory.  
  
```{r}
df_dugongs <- ds |> 
  collect()

head(df_dugongs)
```

We can plot the data to get a better idea of its spatial coverage.  

```{r}
australia <- ne_countries(country = "Australia", returnclass = "sf", 
                          scale = "medium")

ggplot() +
  geom_sf(data = australia) +
  geom_point(data = df_dugongs, aes(decimalLongitude, decimalLatitude), 
             color = "red") +
  theme_bw()
```

We will concentrate our analysis on the last year when data was available. We will add a year column to filter the dataset.  
  
```{r}
df_dugongs <- df_dugongs |> 
  mutate(year = year(eventDate)) |> 
  filter(year == max(year))

head(df_dugongs)
```


```{r}
ggplot() +
  geom_sf(data = australia) +
  geom_point(data = df_dugongs, aes(decimalLongitude, decimalLatitude,
                                    colour = organismQuantityType, 
                                    size = organismQuantity)) +
  lims(y = c(min(df_dugongs$decimalLatitude), max(df_dugongs$decimalLatitude)),
       x = c(min(df_dugongs$decimalLongitude), max(df_dugongs$decimalLongitude))) +
  theme_bw()
```


# Connecting to IMOS THREDDS server
The IMOS THREDDS server can be accessed using the `thredds` package in R. It is also possible to explore the catalog to find the dataset we are interested in, which we show below, but we recommend using the web browser for easier navigation. 
  
```{r connect_thredds}
# We define the URL of the IMOS THREDDS catalog
imos_cat_url <- "https://thredds.aodn.org.au/thredds/catalog/IMOS/catalog.xml"
# We create a CatalogNode object to connect to the IMOS THREDDS catalog
imos_cat <- CatalogNode$new(imos_cat_url, prefix = "thredds")
```

## Exploring the IMOS THREDDS catalog
We can open the catalog in the web browser to explore the datasets available as follows:
  
```{r}
imos_cat$browse()
```
  
Once in the browser, you can change the extension of the catalog URL from `catalog.xml` to `catalog.html`. This way you can explore the catalog in a more user-friendly format.  
  
As mentioned previously, we can also explore the catalog programmatically, but this is more time-consuming. We will check the SST datasets inside the SRS catalog:
  
```{r}
# Listing datasets in the catalog
imos_cat$get_catalogs()
```
  
We can see there are a number of datasets available in the IMOS catalog. We are interested in the `SRS` products, so we can check the datasets available in the `SRS` catalog:
  
```{r}
imos_cat$get_catalogs()[["SRS"]]
```
  
We can see that the `SRS` catalog contains the `SST` dataset we are interested in. There are multiple datasets available inside `SST`, but below we show you the path to the `ghrsst` dataset, which contains the Sea Surface Temperature data we want to access:

```{r}
imos_cat$get_catalogs()[["SRS"]]$get_catalogs()[["SST"]]$
  get_catalogs()[["ghrsst"]]$get_catalogs()[["L3SM-1d"]]$
  get_catalogs()[["day"]]$get_catalogs()[["2023"]]
```
  
We can see that there are 163 files available in the `2023` folder, which contains the SST data for that year. We also get a URL that will allow us to explore the catalog further. We will use this URL to download the SST data for the duration of the 2023 dugong aerial survey.

```{r}
# First we connect to the catalog
srs_cat <- CatalogNode$new(paste0("https://thredds.aodn.org.au/thredds/catalog",
                                  "/IMOS/SRS/SST/ghrsst/L3SM-1d/day/2023/", 
                                  "catalog.xml"),
                           prefix = "thredds")

dugong_dates <- df_dugongs |> 
  distinct(eventDate) |> 
  mutate(date = str_remove_all(as.character(eventDate), "-")) |> 
  pull()

# Now we can get the dataset names available in the catalog
sst_names <- srs_cat$get_dataset_names() |>
  str_subset(paste0(dugong_dates, collapse = "|"))

# Checking we have selected the correct datasets
sst_names
```
  
We have seven datasets available for the duration of the 2023 dugong aerial survey, which is exactly what we wanted. Once again, this method to explore the catalog is more time-consuming, so unless you're familiar with the structure of the catalog, we highlight recommend you use the web browser instead.
  
We now need a complete URL to download the SST data as shown below.

```{r}
# Get base url for HTTPserver access
http_base <- srs_cat$list_services()$http[["base"]]

# Put together full url to download selected files
sst_urls <- sst_names |> 
  map_chr(\(x) paste0("https://thredds.aodn.org.au", http_base,
                      srs_cat$get_datasets()[[x]]$get_url()))

# Checking results
sst_urls
```
  
We are now ready to download the data.  

## Download SST data from IMOS THREDDS server
  
```{r, eval = FALSE}
# We will ensure we have a directory to store the data
if(!dir.exists("data")){
  dir.create("data", recursive = T)
}

# We will download the SST data for the duration of the 2023 dugong aerial 
# survey
# The file names will be named the same as in the catalog
map2(sst_urls, sst_names, 
     \(x, y) curl_download(x, file.path("data/", y)))
```
  
Data download may take a few minutes depending on your internet connection. Once the data is downloaded, we can load it into R for further analysis.
  
# Load SST data
  
```{r}
# We can get a list of the downloaded files
sst_files <- list.files("data", pattern = "GHRSST", full.names = T)
```
  
The dataset contains multiple variables, but we are interested in the `sea_surface_temperature` variable. We can subset the data to keep SST data only and then we will convert it from Kelvin to Celsius.

```{r}
# We can load the SST data using the terra package
sst_dugongs <- rast(sst_files)

# Exploring SST data
sst_dugongs
```



You may also have noticed that there is no information about dates in the raster object. We can extract this information from the `source` attribute of the raster object.

```{r}
# Subsetting SST data
sst_dugongs <- sst_dugongs[[names(sst_dugongs) == "sea_surface_temperature"]]

# Getting dates from the source attribute
date_sst <- sources(sst_dugongs) |> 
  str_extract(pattern = "20231[0-9]{3}") |> 
  as.Date(format = "%Y%m%d")

# Adding dates as time and names to the raster object
time(sst_dugongs) <- date_sst
names(sst_dugongs) <- date_sst

# Transforming from Kelvin to Celsius
sst_dugongs_celsius <- sst_dugongs - 273.15
sst_dugongs_celsius
```
  
# Extracting SST data for points of interest
We can extract SST data for specific points of interest using the `extract` function from the `terra` package. For example, we can extract SST data for a points during the 2023 dugong aerial survey.

```{r}
# Defining points of interest
points_of_interest <- df_dugongs |> 
  distinct(decimalLatitude, decimalLongitude) |> 
  # Converting points to a spatial object
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

# Extracting SST data for points of interest
sst_points <- terra::extract(sst_dugongs_celsius, points_of_interest, xy = T)

# Checking extracted SST data
sst_points
```
  
We can reformat the extracted data to make it easier to read and visualize.
  
```{r}
sst_points <- sst_points |> 
  pivot_longer(cols = !c(ID, x, y), names_to = "date",
               values_to = "sst_celsius") |> 
  mutate(date = as_date(date))

sst_points
```
  

# Calculating mean SST for the duration of the 2023 dugong aerial survey

```{r}
mean_sst <- mean(sst_dugongs, na.rm = TRUE)
mean_sst
```
  
# Plotting SST data
We can create a simple plot of the mean SST using base R.

```{r}
plot(mean_sst)
```
  
We can also create a more informative plot using `ggplot2` and `tidyterra` packages. We will use the `geom_spatraster` function to plot the raster data and add a base map of Australia using the `rnaturalearth` package.

```{r}
# Getting a base map of Australia
aus <- ne_countries(country = "Australia", scale = "medium",
                    returnclass = "sf")
```

Remember that when using `ggplot2`, we create a plot by stacking layers. This means we will need to add the raster data as the first layer and then add the base map as a second layer. We can also add points for the locations we extracted SST data from.  
  
```{r}
ggplot()+
  # Using tidyterra to plot the raster data
  geom_spatraster(data = mean_sst) +
  # Setting the color scale for the SST data
  scale_fill_viridis_c(option = "C", name = "SST (°C)") +
  # Adding points for the locations we extracted SST data from
  geom_sf(data = points_of_interest, color = "darkgreen", size = 3) +
  lims(y = c(min(df_dugongs$decimalLatitude)-1,
             max(df_dugongs$decimalLatitude)+1),
       x = c(min(df_dugongs$decimalLongitude)-1,
             max(df_dugongs$decimalLongitude)+1)) +
  # Adding labels and theme
  labs(title = "Mean Sea Surface Temperature (SST) for 2023 campaign",
       x = "Longitude", y = "Latitude") +
  theme_bw()
```
  
We can also make a plot comparing the mean SST for the duration of the 2023 dugong aerial survey at our points of interest.

```{r}
sst_points |> 
  group_by(date, ID) |> 
  summarise(mean_sst = mean(sst_celsius, na.rm = T)) |> 
  ggplot(aes(x = date, y = mean_sst, color = ID)) +
  geom_line(aes(group = ID)) +
  geom_point() +
  labs(title = "SST at Points of Interest for 2023 campaign",
       x = "Date", y = "SST (°C)", color = "Point ID") +
  theme_bw()
```
  
Finally, we can also create a multipanel plot to visualize the SST data for each day of the 2023 dugong aerial survey at our points of interest.
  
```{r}
ggplot()+
  geom_spatraster(data = sst_dugongs_celsius)+
  scale_fill_viridis_c(option = "C", name = "SST (°C)") +
  geom_sf(data = australia, fill = "#fff2cc", color = "#fff2cc") +
  facet_wrap(~lyr) +
  theme_bw() +
  labs(title = "Sea Surface Temperature (SST) for 2023 campaign") +
  theme(legend.position = "bottom", legend.title.position = "top", 
        axis.text.x = element_text(angle = 45, hjust = 1))
```
  
Remember that you can save any plots as follows:
  
```{r, eval = FALSE}
ggsave("outputs/sst_australia_2023.png", dpi = 300, 
       bg = "white")
```

# Conclusion
In this notebook, we have gone through the basics of how to access IMOS data available via the THREDDS server, specifically the Sea Surface Temperature (SST) data. We have shown how to download the data, load it into R, extract SST data for specific points of interest, and visualize the data using `ggplot2` and `tidyterra`. We have also shown how to calculate the mean SST during the 2023 dugong aerial survey campaign and created multifaceted plots to visualise the data.
