---
title: "Seabird impacts"
date: "2026-08-30"
output: 
  github_document:
    toc: true
    html_preview: true
always_allow_html: true
---

```{r, echo = F}
knitr::opts_chunk$set(fig.path = "figures/") 
```


# Introduction



# Loading required libraries
  
```{r libraries, results = "hide", message = F, warning = F}
library(arrow)
#Data manipulation
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(stringr)
#Plotting
library(ggplot2)
library(tidyterra)
#Base maps
library(rnaturalearth)
#Spatial data
library(sf)
#Gridded data
library(terra)
#Connection to THREDDS
library(thredds)
#Downloading data
library(curl)
```

## Accessing the Dugong Aerial Survey dataset with `arrow`
From the metadata catalogue we obtain the URI and create a connection to the dataset using the `arrow` package. By using the `open_dataset` function, we are able to access the data stored in the server, but we don't load the dataset just yet. Instead we will inspect the contents of this dataset to check if our computer can handle the entire dataset being loaded.

```{r}
ds <- open_dataset(
  "s3://aodn-cloud-optimised/aggregated_seabird_nonqc.parquet")
```

We can check the file structure (i.e., schema) of the dugong dataset. This will give us the name of the columns contained in the dataset, which can help us filter the data before accessing.  

```{r}
ds$schema
```
```{r}
# -38.047884, 145.454860
# -39.564776, 150.471827


seabird_gipps <- ds |> 
 filter(decimalLatitude >= -40 & decimalLatitude <= -38 &
          decimalLongitude >= 145 & decimalLongitude <= 150.5) |> 
  collect()
```

From here, we can see there are `r length(ds$schema)` columns, but we can check how many records in total are included in the dataset as follows:  

```{r}
seabird_gipps |> 
  count(year(eventDate))
```
```{r}
australia <- ne_countries(country = "Australia", scale = "medium",
                          returnclass = "sf")


ggplot()+
  geom_sf(data = australia)+
  geom_point(data = seabird_gipps, 
             aes(decimalLongitude, decimalLatitude, 
                 colour = as.factor(year(eventDate))))+
  lims(x = c(min(seabird_gipps$decimalLongitude),
             max(seabird_gipps$decimalLongitude)),
       y = c(min(seabird_gipps$decimalLatitude),
             max(seabird_gipps$decimalLatitude)))+
  theme_bw()
```


  
